Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-04-14 endofquarterpaper.qmd

@@

The geographic boundaries of the groups change from phase to phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe. This wider distribution may be a factor in weak connections seen in @fig-neighbor-net. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the stronger connectivity seen there. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase. In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.

One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites. The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons. Aggregation sites are the places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994]. It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignroacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose. However, without examining environmental data, this remains only a hypothesis at this time.

The second aim of our study was to determine what these patterns tell us about dynamics of social organization throughout the Aurignacian. The results suggest that the Transitional phase was sparsely populated, with minimal interaction between loosely connected groups. For the Proto-Aurignacian, the results suggest a large expansion in social activity, as well as strongly intra-connected groups with some inter-connectivity present, with geographic distance having the least influence on sign distribution. The Early Aurignacian experiences a large surge in inter-connectivity, as well as intra-connectivity strength, with a great deal of connection between groups and an increase in the correlation with geographic distance. The Evolved Aurignacian experiences a decline in connection strength, although it maintains degrees of inter-connectivity.

These results are consistent with existing data on the environmental and population change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP. This rapid expansion could be what stimulated the emergence of signs as a marker of group membership. Other studies have also found a similar associations in other regions and time period between increases in quantity and diversity of art, and increases in population density [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992].

Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.

This study adds to our knowledge of social networks within the Aurignacian period, as our results suggest dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our network analysis shows evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We hope that dividing the data up into smaller time periods serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023]. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].

One possible limitation is the potential ambiguity that arises with some of the sign types. While @Von_Petzinger_2009 argues in favor of Aurignacian geometric signs as symbolic expression, potential ambiguity still remains over whether function is consistent across all sign type occurrences, particularly in regards to the simpler sign types composed from fewer elements. We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023]. Potential ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains information on the specific order and patterns in which signs occurred.

@@

This study both adds to our understanding of Aurignacian social dynamics and suggests potential new directions for future research. The observed social dynamics produced by our methods align with the existing data on the environmental and population conditions of the Upper Paleolithic. We feel that our research opens up new opportunities for the application of these methods to data on artwork variation in other regions and time periods. Our research also suggests an earlier emergence of dynamic inter-connectivity than previously predicted, as we see weaker inter-connections beginning to emerge by 41 k BP, and stronger inter-connections by 38 k BP. Future research directions might involve the analysis of environmental conditions and their interactions with social dynamics, in order to determine whether some of the sites that we have pointed to as potential aggregation sites function as such. Another potential future research direction would be the usage of the more specific data released in SignBase 2.0 in order to counter potential sign type ambiguity.

>>>>

## Grouping Dynamics Across the Four Phases

@@

The geographic boundaries of the groups change from phase to phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe. This wider distribution may be a factor in weak connections seen in @fig-neighbor-net. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there. This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase. In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.

One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites. The aggregation/dispersion pattern is first defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994]. It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.

## Dynamics of Social Organization

The second aim was to determine what these patterns tell us about dynamics of social organization throughout the Aurignacian. The results suggest that the Transitional phase was sparsely populated, with minimal interaction between loosely connected groups. For the Proto-Aurignacian, the results suggest a large expansion in social interaction within groups, as well as some degree of interaction between groups, with geographic distance having the least influence on sign distribution [@tbl-mantel-table]. The Early Aurignacian experiences a large surge in interaction between groups, as well as interaction within groups, and an increase in the correlation with geographic distance. The Evolved Aurignacian experiences a decline in between-group interaction, although it maintains within-group interaction.

These results are consistent with previous work on the environmental and population change during these phases. From 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans [@Shao_Wegener_Klein_Schmidt_Weniger_2024]. This was followed by rapid expansion from 43.25 - 41 k BP (Transitional/Proto), with the maximum extent of Proto/Early Aurignacian settlement being reached around 41 k BP. This rapid population expansion could be what stimulated the emergence of signs as a marker of group membership. Other studies have also found similar associations in other regions and time periods between increases in quantity and diversity of art, and increases in population density \[@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992\].

Similarly, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 (HS4) at the very beginning of the Early Aurignacian resulted in a expansion of people into new environmental niches, which they hypothesize was associated with an expansion of social networks both within and between populations. This correlates with our results which show a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased sign type correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. This geographic expansion may explain our results, where we find weaker, more dispersed connections in the Evolved Aurignacian.

\<--

cite signbase 2.0–\>.

@@

This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].

One limitation is the potential ambiguity that arises with some of the sign types. While @Von_Petzinger_2009 argues in favor of Aurignacian geometric signs as symbolic expression, potential ambiguity still remains over whether function is consistent across all sign type occurrences, particularly in regards to the simpler sign types composed from fewer elements. We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023]. This ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains more detailed information on the specific order and patterns in which signs occurred on individual objects

\<--

cite signbase 2.0–\>.
